Iterative Monomorphisation

### 2025-02-13 iterative_monomorphisation.tex
tune

We described a procedure for iteratively instantiating polymorphic types to
produce monomorphic problems as a preprocessor. Our primary motivation was to
improve the success rate of Zipperposition and its monomorphic E backend, and
indeed our evaluation shows a clear improvement. We also saw that even with
automatic provers that support polymorphism, iterative monomorphisation is a
better alternative in practice.

\looseness=-1
We see the following avenues for future work. First, iterative monomorphisation
blindly enumerates candidate instantiations, without exploiting any knowledge
about the logical structure of the formulas in which symbols occur. For
example, a lemma $\sym{p}\langle\alpha\rangle$ cannot be used to prove the
conjecture $\lnot \sym{p}\langle\ty{nat}\rangle$ because of the incompatible
polarities, but our procedure instantiates $\alpha$ with $\ty{nat}$ regardless.
Second, some automatic provers as well as tools such as Sledgehammer include
relevance filters that heuristically select a subset of the available axioms;
filters such as MePo~\cite{meng-paulson-2009} and SInE
~\cite{hoder-voronkov-2011} are iterative and could be interleaved with
monomorphisation. Third, although one would expect native implementations of
polymorphism to outperform any preprocessor, this is not the case now, therefore indicating that there is considerable room for improvement on the native front.

>>>>

We described an algorithm for iteratively instantiating polymorphic types to
produce monomorphic problems as a preprocessor. Our primary motivation was to
improve the success rate of Zipperposition and its monomorphic E backend, and
indeed our evaluation shows a clear improvement. We also saw that even with
automatic provers that support polymorphism, iterative monomorphisation is a
better alternative in practice.

\looseness=-1
We see the following avenues for future work. First, iterative monomorphisation
blindly enumerates candidate instantiations, without exploiting any knowledge
about the logical structure of the formulas in which symbols occur. For
example, a lemma $\sym{p}\langle\alpha\rangle$ cannot be used to prove the
conjecture $\lnot \sym{p}\langle\ty{nat}\rangle$ because of the incompatible
polarities, but our algorithm instantiates $\alpha$ with $\ty{nat}$ regardless.
Second, some automatic provers as well as tools such as Sledgehammer include
relevance filters that heuristically select a subset of the available axioms;
filters such as MePo~\cite{meng-paulson-2009} and SInE
~\cite{hoder-voronkov-2011} are iterative and could be interleaved with
monomorphisation. Third, although one would expect native implementations of
polymorphism to outperform any preprocessor, this is not the case now, therefore indicating that there is considerable room for improvement on the native front.
